Using simple linear models with observational data may make sense as phenological responses to most cues are expected to be linear except at the extremes. Natural conditions often see only a small slice of the range of values of each cue that are possible---and those values often appear to be in the middle---potentially linear---range \citep{gauzere2017,ettinger2020}. Further, given that interactions between cues are difficult to estimate---more so when those cues are highly correlated in nature---focusing on the main effects of cues may provide more robust estimates from observational data  \citep[main effects integrate over any interactions and require lower statistical power to robustly estimate compared to interactions,][]{gelman2006}.


Measuring the three major cues and understanding how they will interactively determine current and future phenology is hard for several major reasons. First, these cues vary across species, and possibly within species across the range  \citep{vitasse2009,harrington2015}. Second, the cues often interact: one cue may compensate for another cue \citep{Chuine2000}. In practice this means studies that are not designed to tease apart these cues experimentally (e.g., observational studies) may find statistical evidence for only one cue, because it masks one or more other cues (ADDCITES). For example, a plant that has not received enough chilling would generally require more forcing and thus leafout later, which could be indistinguishable (in observational data) from a photoperiod requirement that has not been met. \\

To some extent these interactions between cues have meant that---for many locations in most years---researchers have not been challenged to measure cues beyond forcing to explain advances and make accurate near-term predictions. In many locations if chilling cues are met and forcing produces budburst at an acceptably long photoperiod---then the main cue needed for predictions is forcing alone. In such cases, simple linear models of spring warming versus spring phenology may suffice \citep[e.g.,]{Ellwood2012}. But, increasing evidence, in some systems in some years, suggests that these chilling and photoperiod may play an important role in spring phenology today, with that importance growing in the future \citep{chuine2016,gauzere2019}. If rising evidence suggests these cues are critical to understand current responses and predict future non-linearities, then how do we integrate them more into phenological research and forecasting? \\% Thus, the cues in observational data---even long-term records---can be very hard to measure. 
% Dan thought the text worked through here...

These cues may create non-linear responses with warming in two major ways. First, as discussed above, each cue non-linear ... Secondly, interactions between cues


Improved chilling models for forest trees (and horticultural trees) may come from an improved understanding of dormancy at the physiological level. \\ % And there is so much we don't know about how chilling works and interacts with forcing (sequential model, parallel models etc.)
% Nacho sayd: could we say this in a positive note? i.e. what should be done to make more robust models? 
% End of Cat's concerns

Similarly the complexity of chilling could produce myriad non-linearities.  

Recent observations of declining responses to warming in long-term data suggest cues beyond forcing may be playing a role \citep{fu2015} and highlight the major need for non-linear models based on interactive cues. To this aim, a number of recent papers have attempted to identify chilling or photoperiod cues in long-term data. Correlations between these predictors, however, are almost always too high for any useful analysis \citep{sarahailene2020,ettinger2020}. 