\documentclass[11pt,letter]{article}
\usepackage[top=1.00in, bottom=1.0in, left=1.1in, right=1.1in]{geometry}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}

\def\labelitemi{--}
\parindent=0pt

\begin{document}
\bibliographystyle{/Users/Lizzie/Documents/EndnoteRelated/Bibtex/styles/besjournals}
\renewcommand{\refname}{\CHead{}}


{\bf Titles}

Chilling dominates tree budburst in controlled climate experiments, but not in the great outdoors
Chilling outweighs photoperiod and forcing cues in temperate trees in experiments, but not in natural systems


\begin{abstract}
Decades of research on woody species highlight how three major cues shape spring phenological events (e.g., budburst and leafout): forcing (warm temperatures, generally occurring in the late winter and early spring), daylength (photoperiod) and chilling (cool temperatures, generally occurring in the fall and late winter). How pervasive these cues are and whether some species are effectively governed by only one or two cues is a critical area of climate change biology research, as it would shape how complex responses to warming will be. Here we use a global meta-analysis of all published growth chamber studies to test for the relative effects of these three major cues across XX species. We find they almost all show these cues, making climate change responses complex. 
\end{abstract}

{
\section{Outline so far...}

For decades, plant phenology has been one of the most reported and consistent biological imprints of climate change: many temperate plants have shifted their leafing and flowering earlier with rising temperatures. Understanding such shifts is important as phenology shapes a suite of ecosystem services, including pollination and carbon sequestration, and scales up to impact projections of climate change itself.  As research interest in phenology has progressed, discrepancies and uncertainties in our understanding have emerged, however. Though responses to warming are consistent on average, they show high variation across species and sites. Furthermore, long-term observational data provide increasing evidence of weakening sensitivities of phenology to temperature (i.e. some of the most-studied species suggest the long-term trend towards ever-earlier springs may be stalling in parts of Europe). 
\par Three major cues shape spring phenology. Chilling (cool temperatures, generally occurring in the fall and late winter). Forcing (warm temperatures, generally occurring in the late winter and early spring), daylength (photoperiod) .  A fundamental challenge of understanding the relative roles of these three cues is that they are often strongly correlated. During the transition from winter to spring at many temperate latitudes, air temperatures increase (i.e. forcing increases) at the same time that daylengths increase. 

\par Chamber experiments: their value --  often attempt to break this correlation to reveal mechanistic links between environmental conditions and budburst date. Chamber studies to date have constrasting effects (Zohner, Laube, Basler, Caffara etc.)
\begin{enumerate}

\item Meta-analysis to the rescue!
\begin{enumerate}
\item Lit review of all woody plant spring phenology (total number of studies \& species, ref map, span of years)
\item Used only studies for which we could figure out forcing, photo and chill, chill we often calculated ourselves as so rarely reported (total number of studies, species in model)
\item To estimate the cues we used a Bayesian hierachical model 
\item This partially pools for a robust overall effect, and for robust effects for species with lots of data (\emph{Fagsyl, Betpen}) but pools towards the mean for species with fewer data ... (mention species complex here or put in caption)
\end{enumerate}
\item Short paragraph: Our results show that budburst phenology is determined by forcing, chilling, and photoperiod -- all three cues are important and all three advance budburst. Consistency across species -- fairly consistent with some variation in forcing, and then chilling, not much for photoperiod. 
\item Temperature, which is radically altered by climate change, was most important -- chilling and forcing show large effects on budburst
\begin{enumerate}
\item Chilling is the strongest strongset and most consistent cue (ref Laube and anyone else?)
\item Then forcing, consisent with experimental studies (CITES), observational (CITES)
\item Chilling is rarely manipulated directly, thus we had to calculate most of the chilling (impossible to provide estimates with only experimental chilling... ref supp heat maps)
\item Weinberger methods is most common for chilling and this is not a super way to measure it.
\item How you measure chilling matters a bit ... Utah vs. Chill portions
\end{enumerate}
\item Photoperiod
\begin{enumerate}
\item Photoperiod ... very consistent across species, suggesting all species do cue to it (ref Zohner, Caffara, Flynn ??)
\item The magnitude of photoperiod effects varies with latitude, with lower source latitudes generally having earlier budburst. Say provenance (population). 
\end{enumerate}
\item We did not estimate interactions. Why? ..  Add in node to process based models here. 
\begin{enumerate}
\item Very few studies actually design experiments to test for interactions, so there is little to build on
\item The few studies that do interactions often use the weinberger method, which seems a little weird based on our results.
\item They're hard for a couple reasons: need more reps, and photothermoperiodicity.
\item And! We cannot fully disentangle forcing vs. chilling conditions. (Chuine et al. 2016 GCB).
\item Our results average over interactive effects. 
\end{enumerate}
\item One paragraph: A simple interpretation of our model -- especially its chilling and photo effects -- predicts declining sensitivities in long-term data with climate change. This is because even though forcing increases, chilling is expected to decreases and photoperiods should get shorter -- both predicting delays, and thus an overall muted effect of temperature-only.  (Ref exp conditions forecasting figure.) But how do experimental temperature and photoperiod compare to predicted ones in nature? (Ref experimental conditions forecasting figure)
\item But how do the conditions overlap with natural conditions? (PEP + experimental data figures)
\begin{enumerate}
\item Forcing isn't bad
\item Experimental chilling is generally lower than field chilling
\item Photoperiod differences are very big in experiments
\item Declining sensitivities in PEP data (need to check)
\end{enumerate}
\item Forecasting with these semi-real data, however, do not predict a decline in sensitivity given the moderate amounts of warming already seen, instead they a suggest general advance of budburst until extremely high warming (ref. forecasting figure with PEP-based data)
\begin{enumerate}
\item Chilling often increases with small amounts of warming in some sites
\item Even if warming only happens in the winter, it takes a lot of warming to see a delay due to decreased chilling
\item At higher warming do see a leveling off or delay due to decreased chilling at some sites
\item Depends a lot on local climate... We also find that patterns of advancment with warming vary considerably depending on the current/background climate (e.g. how much advancement will continue with warming depends on how much chilling is currently experienced and whether that will increase or decrease with warming.)
\item (Compare advances in our models to PEP725 data?)
\item Photoperiod effects are minimal, even for \emph{Fagus}
\end{enumerate}
\item So why is PEP725 showing declining sensitivities?
\begin{enumerate}
\item Our results suggest few sites with delays before 3-4 degrees warming (CHECK)... and Germany has warmed X amount
\item Speeding up a biological process given sampling time resolution could lead to declining estimates of sensitivites, even if unchange
\item Say something about what to do about this and how to figure out if this is the issue or it's cues. 
\end{enumerate}
\item Our results suggest most or all studied species are responsive to these three cues
\begin{enumerate}
\item Our results are only for one region, but highlight how critical accurate forecasts of shifts in forcing and chilling will be at local scales
\item To do this, we desperately need to better understand chilling (dormancy release) so that we can predict it in the future (maybe say need better models for chilling across species). 
\item Alongside this, we need more fundamental understanding of interactive cues, which requires larger studies across diverse species. Our results include these complexities but a finer understanding is needed in locations where cues do not change in concert.
\item These complexities are unlikely to alter our fundamental predictions of an increasing advance for many temperate trees in the future, even those with strong chilling or forcing cues (ref Gauzere) [Alt: An imrproved understanding of interactive cues, however, is unlikely to alter our fundamental predictions of an increasing advance for many temperate trees in the future, even those with strong chilling or forcing cues (ref Gauzere), unless cues are changing very asynchronously.]
\end{enumerate}
\end{enumerate}

\bf Main text figures:}

\begin{enumerate}
\item $\mu$ plots
\item  $\mu$ forecasting figures: spring x winter warming -- PEP climate range and experimental climate range
\item Species forecasting with PEP data: \emph{Betula, Fagus} ... need to think on which ones to use (x sites x species focus etc). ... Maybe show photoperiod one?
\item PEP data figure with environmental conditions: as in Cat's figure + OSPREE data + maybe foercasting (at 2C or such?)
\end{enumerate}


{\bf Supplemental figures/tables:}
\begin{enumerate}
\item Map of study locations, shading or symbol coding for number of cues (Lizzie)
\item Map of species forecasting to justify sites
\item Tables, yes.
\item Heat maps for the main data, including by actual study design and by calculated chilling (our calculations)
\item Photoperiod x latitude effects figure
\end{enumerate}

\section{Reference list}

A few categories:\\

Papers about contrasting results over what cues matter from growth chamber studies: \cite{Basler:2012,Basler:2014aa,Caffarra:2011qf,Caffarra:2011a,Caffarra:2011b,Heide:2005aa,koerner2010b,Laube:2014a,vitasse2013,zohner2016}. Get Nanninga \emph{et al.} 2017: 'Increased exposure to chilling advances the time to budburst in North American tree species' and maybe Malyshev \emph{et al.} 2018 `Temporal photoperiod sensitivity and forcing requirements for budburst in temperate tree seedlings.'\\

Papers about declining sensitivities (Ailene will update this list): \cite{Rutishauser:2008fu,fu2015}. Also look for a Wang \emph{et al.} article `Impacts of global warming on phenology of spring leaf unfolding remain stable in the long run.' Vitasse paper on declining variation across elevation gradient. See \cite{yu2010}, but this is not temperate trees. \\

Papers about chilling units paper (Lizzie gets a list): Fu 2012 from OSPREE. \cite{harrington2015}\cite{lued2011,Luedeling:2011qe,Luedeling2013AgFM}\\

\bibliography{..//..//refs/ospreebibplus.bib}


\end{document}


\begin{enumerate}
\end{enumerate}