\documentclass[11pt, a4paper]{article}
\usepackage[top=0.75in, bottom=0.75in, left=1in, right=1in]{geometry}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{gensymb}
\begin{footnotesize}

\begin{document}{}
\bibliographystyle{..//..//..//..//refs/bibstyles/Science.bst}

\pagenumbering{gobble}
\setlength\parindent{0pt}


\title{Response to Reviewers}
\emph{Reviewer Comments are in italics.} Author responses are in plain text.\\

 \emph{\bold{Reviewer \#1 (Remarks to the Author)}}\\

\emph{The relative importance of forcing, chilling and photoperiod as cues for budburst is a fascinating one, with clear implications for predicting how species will respond to climate change. Here the authors leverage an exceptional dataset arising from experimental studies using sophisticated statistical analyses and arrive at the surprising conclusion that plants are generally more sensitive to chilling than forcing. I think this study has the potential to make a really valuable contribution that will be of broad interest to readers of this journal. However, I have quite a lot of criticisms/concerns of the study as it stands.}\\

\par We thank the reviewer for the recognition that the OSPREE dataset is exceptional and that our study can make a valuable contribution to climate change researchers. We have revised the original manuscript substantially to address the reviewer's concerns, as detailed below. \\

\emph{(1) Models: The STAN modelling approach is sophisticated but I think the model is rather incomplete and this could affect the inferences that are reached. For instance why aren't terms included to allow the intercepts and slopes to vary across studies within species?} \\

% EMW: Also, I think we should mv up that we made a change and move more of the discussion down... otherwise I think with some tweaking to the main text this is good. 
\par We completely agree with the reviewer that, ideally, our models would account for both variation in budburst responses among species and among studies. This was an issue we worked extensively on, but we did not clearly present our approach in our initial submission and have revised the text. Our aim in building the OSPREE database and using Bayesian models that can allow complex models to converge was to cleanly separate out species from study effects. During our data and model development, however, we found studies were generally confounded with species. This is a common problem in this area of research and other meta-analyses have faced similar issues \citep[e.g.,][, faced the same issue and handled it similarly in consultation with statistician Andrew Gelman]{kharouba2018}. Without strong priors to help differentiate what variation should load onto the study versus species, it is difficult to fit both variables. Indeed, we evaluated models that included studies within species and studies crossed with species but both models were found to be unstable (e.g., estimates varied widely across model runs, chains did not converge). \\

To address this we developed a model that only included species for which we had multiple treatments across multiple studies. We combined species found in only one study into ``complexes'' at the level of genera, such that each taxonomic unit we use in our model occurs across multiple studies (and treatments). Thus our taxonomic units of analysis are ``species complexes,'' which are either species represented in $>1$ dataset or complexes combining multiple species within a genus that are each singly represented in the dataset. Species represented in only one dataset with no con-generics in other datasets were excluded from most of our analyses. We do include a model fit to all species and report these results in the supplement (Tables S3 and S4); we refer to this model as the `all species' model.\\

% EMW: I would also add that we now focus on this model and present the larger model---which gave similar results---only paranthetically. 
We realize now we should have made these issues more clear. We now write on Lines 62-67 in the main manuscript:
\begin{quote}
`Some species are often only represented in one dataset in the OSPREE database, making it impossible to statistically differentiate between species, study, and treatment effects for these taxa. To address this, we combined species found in only one study into ``complexes'' at the level of genera---such that each taxonomic unit we use in our model occurs across multiple studies (and treatments, see the \emph{The Observed Spring Phenology Responses in Experimental Environments (OSPREE) database} section in the Supplemental Materials for details.)
\end{quote}

\emph{Also, I would have thought there is very likely a geographic effect on the effects, and I suggest that you test whether the results are sensitive to inclusion of a spatial random term across which slopes and intercepts vary.}

% EMW: Again, we want the fact that we do deal with this UP FRONT for the editor to see ....  also, I thought we would just do NA vs Europe -- I think that would be okay, but fine if you have done it will all continents...
\par The reviewer makes an excellent point that budburst responses to temperature and photoperiod may vary due to the spatial location of studies or the geographic origin of plant material. In particular, budburst responses are expected to vary by latitude \citep{saikkonen2012,zohner2016,gauzere2017}. We address variation in space primarily by including a latitude model, which we now discuss on lines XX in the main manuscript.  We focus on the spatial effect of latitude because there is strong evidence that there can be latitudinal differences in budburst responses, for example via interactions between latitude and photoperiod sensitivity and interactions with chilling responses (XX CITATIONS). In this new version, we have also added a model that includes continent. This new continent model includes photoperiod, forcing, and a random effect of continent on the intercept, as well as species-level random effects (as in the main budburst model). This new  model is described on page X in the supplement, with results summarized in Table SX. Note that we do not include chilling in this new model because the climate data used to estimate chilling is only available for two continents (North America and Europe). We find that variation by continent XXXX% Yes! The more analyses we can add to bolster our results, the stronger our reply is. Remember we only have chilling for NA and Europe so I think our options are this as a fixed effect or a model of bb ~ photo + force with continent on the intercept (latter seems better to me, include just in supp). 

\emph{(2) Meta-analysis: The analysis is described as a meta-analysis, but falls short of being a formal meta-analysis as it seems as though measurement error in the response variable is not incorporated. This should be straightforward to incorporate and I was surprised that it hadn't been given the complexity of the analyses. Also, please report the extent to which the approach followed recommendations made in the PRISMA checklist }\\

\par This is a good point; we now cite the PRISMA checklist in the Supplement (page 1, first paragraph) and include the full checklist in reference to our study as Appendix 1. Many of the checklist items are done by our data publication, and we feel this addition greatly strengthens the paper and appreciate the reviewer suggesting it.\\

\emph{(3) Methods: The methods seem to be missing from the main ms, and I kept flicking forward to consult a section that does not exist. I thought the Nature letter format does allow a methods section and I found it really to the detriment of the readability of the ms that there wasn't one.}\\

\par We appreciate the reviewer's concern that methods are not easy to find; this was a concern of multiple reviewers so clearly something we needed to improve. We have worked to now more clearly embed key methods in the text. We now have included a more full model description and the model equation (Lines 66- 74), and added a new figure to summarize the experimental methods of studies in our meta-analysis (Figure 1).

\par We also provide a separate Methods section with full details of the data and are analyses. This section will be available online, should our manuscript be accepted for publication in \emph{Nature Climate Change}, following the journal's current requirements.\\

\emph{(4) Chilling, forcing and photoperiod: In order for a reader to reach a conclusion about the robustness of the inferences it is vital that the method for quantifying cues is easily understandable. Currently in the main ms it is not (last paragraph of page 3). For instance, we are told the minimum temperature for chilling but not maximum, we are not informed as to when the chilling and forcing periods are and no discussion is given as to how the effect of photoperiod is modeled. It's also unclear in the main ms what a `standard unit' (I see it is described in the supplement) is and this leaves the reader disconnected with what the analyses are doing. A simple remedy for this would be to include a schematic (as figure 1) that identifies the information used to quantify each cue and relate it to the response. In general the main ms does a very poor job of explaining what was done (the data used, how cues were inferred and vital details about what the models were estimating), instead referring the reader repeatedly to supplementary materials. While the supplementary materials are generally good I still felt disconnected from the data and how the cues were actually quantified. This could be addressed by taking some example datasets and working through in detail how the different metrics were calculated. Without knowing what was done I find it very hard to judge whether the main conclusions are robust.}\\

\par We thank the reviewer for pointing out that important details on the chilling, forcing, and photoperiod estimates used in our analysis were unclear in the original version. We have added a new schematic figure (the new Figure 1), as the reviewer suggested, which we hope clarifies how the chilling, forcing, and photoperiod estimates were obtained from the original studies for use in our meta-analytical work. We have added substantial new text related to details on chilling, forcing, and photoperiod to the main text in the following locations: 


\par We also provide a full description of the Utah model, including the upper and lower thresholds for chilling, in the Methods section, Table XX.\\

\emph{(5) Chilling: I think it's important to know whether the inferences are robust to an alternative model of chilling, e.g., the sequential model that is widely used. From the supplementary materials it is clear that some effort has been made to consider alternatives (chilling portions) but given this analysis underlies the main conclusion of the paper I'd like to see alternative hypotheses considered.}\\

\par This is an excellent point as our earlier version did not adequately compare results with the Utah model to results with the other chilling model we evaluated, Chill Portions. Now we more clearly compare them in the main text (lines xx-xx), where we write: ``We found that applying a different chilling model did not strongly affect our estimates (i.e., 95\% credible intervals of estimates for chilling, photoperiod, and forcing overlapped for models using Utah and chill portions with standardized predictors, Table S4)."  \\

\par We would be happy to add other estimates of chill units beyond the widely used Utah and Dynamic models \citep[e.g.,]{Luedeling2009}, at the reviewer's request. Parallel versus sequential models \citep[as used in][]{fu2015,chuine2016} estimate chilling and forcing effects through a process-based modeling approach and---to our knowledge---are always applied to data with variable temperatures across those periods (e.g., in natural field conditions). Based on this understanding, and in consultation with Isabelle Chuine about applying these models to the controlled environment studies we review here, we do not see a good method for testing the sequential model (or parallel, alternating etc.) on these data. If the reviewer has specific modeling suggestions or other chilling model he would like applied we would be happy to test them. We also hope Figure 1 may clarify this. \\

\emph{(6) Estimates: It is surprising to see point estimates repeatedly reported throughout the ms without 95\% credible intervals, this needs to be rectified. Also at present there is no formal test of whether the chilling response is significantly stronger than the forcing response, though this would be easy to do using the posteriors.}\\

\par We thank the reviewer for this suggestion and have added 95\% uncertainty intervals to all estimates presented in the main text (e.g., Lines 79-91).  In figures, we show 50\% uncertainty intervals to focus on the most likely estimates (we can change this upon request), and in the supplemental tables we present both 50\% and 95\% uncertainty intervals.  \\

\emph{Statistical artefact with linear regression (Page 5): That application of linear regression to data arising from a growing degree model can lead to biased estimates is a fascinating insight. However, in the supplementary materials it is not clear to me how the temperature sensitivity window for linear regression (for B. pendula or the simulations) is calculated/defined. 
How much can the issue of an advancing period of sensitivity be addressed by allowing the sliding window to shift over time? This issue is discussed in Simmonds, E. G., Cole, E. F., \& Sheldon, B. C. (2019). Cue identification in phenology: a case study of the predictive performance of current statistical tools. Journal of Animal Ecology.}\\

\par We thank the reviewer for highlighting the need for additional methodological details, and for this insightful question about the shifting window approach. The window we used was 1 March through 30 April, which we are now stated in the online Methods, section \emph{Applying our model to Central European data}, (Page 6, paragraph 2).
\par To address the reviewers question, we applied the shifting window approach to the same dataset and compared the temperature sensitivities  pre-and post- warming. These sensitivities show the same pattern: higher sensitivity pre-warming compared with post-warming. We have added these analyses to a new section in the ``Methods and Supplemental Materials" entitled \emph{Applying the sliding windows approach to Central European data}.\\

\par \emph{Minor comments}\\

\emph{Page 2. I suggest changing 'high unexplained variation across' to 'substantial variation among'.}\\

\par  We thank the reviewer for this suggestion and have made the recommended change (now Lines 22-23).\\
 
\emph{Page 3. All three cues are not generally correlated in longitudinal studies; photoperiod and forcing are, but neither is usually very correlated with chilling.}\\

We thank the reviewer for pointing out that our writing was not clear in this section. In our earlier version of the manuscript, we mentioned correlations between cues but did not specify clearly whether we meant correlations across space or time, nor were we clear about the scale or window at which these cues can be correlated. Though the reviewer states that chilling is often not correlated longitudinally, we have found that it can be correlated (ADD FIGURE OR JUST r/r2 showinging this?). In addition, chilling and forcing are frequently negatively correlated in space (sites that have experience high chilling tend to have low forcing, ADD FIGURE OR JUST r/r2 showinging this). In this new version of the manuscript, we have clarified our writing, which now says (Lines XX):
\begin{quote}
....
\end{quote}

\emph{Page 3. Last sentence of paragraph 2. This is hyperbole. The mean is not expected to shift far beyond historical bounds, though the extremes clearly will.}\\

The reviewer appears to have concerns about the following phrase from the previous version of our manuscript: ``... continued warming pushes climate into environmental regimes far beyond historical bounds.'' We thank the reviewer for highlighting this phrase, which is a bit vague in its reference to `environmental regimes' and for which we clearly should have included citations to support. We now cite the IPCC forecasts that climate change is expected to push temperatures to XXX, well beyond the historical bounds. We adjusted the sentence to be more clear and have added references so that it now says (Lines XX): ``Resolving these discrepancies is critical to accurate predictions of spring phenology, especially as continued climate change will yield warmer temperatures than has been experienced in at least 150 years and warming will occur at a rapid pace (Ohlemuller et al 2006,Williams et al 2007, Williams et al 2007b,IPCC 2013, Xu et al 2018, FIX REFS!).\\

\emph{Page 3. Fourth line from bottom. Is interactions the correct term?}\\

We apologize that this was unclear. We have now adjusted this sentence (lines xx-xx), which now reads ``Our model averages over interactive effects of predictors, as it includes only main effects; it estimates both species-level responses...''\\

 \emph{Reviewer \#2 (Remarks to the Author):}\\

\emph{Spring leaf-out phenology plays a key role in terrestrial carbon and water flux, but the underlying processes are still unclear, especially how the environmental cues, including chilling, photoperiod, and spring warm temperatures, interact and determine the leaf-out processes is still unclear, although most of the phenologist agreed that these three cues are all important. Therefore, quantify the relative importance are valuable and might be important for the phenology modeling and dynamics vegetation models. I carefully read this meta-analysis and found this is an interesting study, but I'm wondering, given the results were reliable, whether the meta-analysis results across experimental studies could reflect the natural plants' response? Or could we rely these experimental results that may inaccurate reflect underlying mechanisms? Because, according to the author (E.M. Wolkovich) previous study, the phenology under warming experiments could not reflect the natural observations (Wolkovich et al, 2012 nature, warming experiments underpredict plant phenological responses to climate change), which might arise from complex interactions among multiple drivers and remediable artefacts in the experiments that result in lower irradiance and drier soils.}\\

\par We thank the reviewer for pointing out that there are limits to the controlled environment studies synthesized in our meta-analysis. We agree that there are many ways in which experimental conditions differ from observational conditions and we were not clear about this in our previous version. The reviewer's comment also highlights that the previous version did not adequately describe the experimental methods used in controlled environment studies. \\
\par To address these comments, we have added a new figure (Figure 1) to show the experimental design of the controlled environment studies in our synthesis. In this new version, we  also discuss these issues in the following places:
\begin{enumerate}
\item  Lines 42-47: ``In contrast to observational studies, controlled environment experiments can break down correlations between the cues. Note that controlled environment studies differ from field warming experiments, which are designed to test higher temperatures in natural conditions, because they are designed to create conditions less often seen in nature, but which can critically help researchers identify cues. These experiments---most often conducted in growth chambers or similar systems to control temperature and light (Fig.1)---have been conducted for decades and their accurate representation of phenology in the natural world has been tested (Vitasse and Basler 2014). "
\item  Supplement Pages 5-6, \emph{Applying our model to Central European data}: ``We also wished to understand how our findings may apply to conditions more commonly found in nature, where conditions often vary dramatically from those applied in controlled environment experiments. For example, very low amounts of chilling can be applied in experiments compared to the natural chilling found in many temperate areas (Fig. S5). Additionally, chilling temperature and total chilling are more correlated in nature than in experimental conditions (Fig. S5). Further, given the importance of chilling and forcing combined with the fact that seasons do not always warm uniformly with climate change (17; 18), we also wished to understand how warming in the winter, spring, or both seasons would shift budburst timing.''
\end{enumerate}

\par  We hope that these changes clarify that the controlled environment studies in the present paper differ from  the field-based warming experiments in Wolkovich et al 2012. Controlled environment studies are typically designed to tease apart the role of different cues (chilling, photoperiod, forcing), rather than to replicate current or future natural conditions (the warming studies synthesized in Wolkovich et al 2012  are typically designed to replicate natural conditions). \\

\emph{Furthermore, I'm not convinced that the chilling overweight forcing, and the effect of chilling, photoperiod and forcing might be quantified across more than 200 species based on the various manipulative experiments and MCMC-based Bayesian method, especially considering most of these experimental studies conducted only one year or less than 3 years. The main reasons come from: 1) most of these experimental studies conducted with very different settings, such as using saplings vs. mature tree?s cuttings, how the ontogenetic effects play a role or impacts the results? Arbitrary controls in lights/photoperiod length/intensity vs. greenhouse natural light; in addition, for many experimental studies, the temperature and photoperiod were set under extreme climates. I would say this is a response to extreme climate. All of these factors might substantially affect the results.}\\

\par We agree with the reviewer that there are many factors, not included in our main model, that could explain variation in budburst responses to chilling, photoperiod, and forcing (e.g., ontogeny). In this version we have tried to more strongly indicate this, as well as state that separating chilling from forcing requires more physiological research. For example, please see Lines XXX, where we state:
``''
\par In addition, in this new version we include a new model to directly test for ontogenetic effects on budburst by adding a predictor of "life stage" (juvenile vs. adult) to the main budburst model. We found that material from juvenile trees (seedlings or saplings) burst, on average, 3.6 days later than material from adult trees. We have added this model and a table summarizing its results to the supplemental materials. 
\par We believe that there are many additional potential questions and avenues of research to better understand how woody plant phenology responds to different cues, at varying life stages, and in different contexts. Our database and analyses provide a critical step in pushing this work forward by providing the first comprehensive meta-analysis quantifying responses to chilling, photoperiod, and forcing; it is our hope that our analyses, as well as the freely available OSPREE database, stimulates future experiments and analyses to test additional hypotheses. \\

\emph{2) the interact between chilling, photoperiod and forcing is complicated, and there are still unclear in many important facts. For example, the temperature thresholds of chilling and forcing estimation, and its species-specific values, are largely unknown. For some boreal or alps plants, they may budburst even when air temperature around freezing points, but the temperate trees are still dormancy even air T > 15 degree; the correlations between eco- and endo-dormancy, corresponding the chilling and forcing, whether they are a parallel or a sequential pattern between chilling and forcing? When/how the photoperiod plays its role during the two phase dormancy? Once the endo-dormancy break, continuous chilling accumulation, for example a cold span during spring, is still active? Or entirely depending on the forcing? All these questions are still not figured out; 3) except chilling, forcing and photoperiod, other cues are also involved with the leaf-out processes, for example air humidity, see Laube et al, 2014 (but recently, Zohner et al, 2019 New phytologist deny this effect) and soil moisture and snow cover. Under manipulative conditions, these effect might be largely ignored as argued in Wolkovich et al, 2012 as well.}\\

\par We completely agree with the reviewer that the interactions between chilling, photoperiod and forcing is complicated. We now discuss this on lines xx-xx, where we write: ``Linking such short-term controlled experiments to natural conditions robustly will require more efforts to understand the complex interactions between chilling, forcing, and photoperiod that we were not able to quantify in this meta-analysis. Most experimental studies do not test for interactions between all three cues (Table \ref{}). Further, many additional factors can affect phenological responses, including ontogeny \citep[Table \ref{tab:stage},][]{vitasse2013ont}, latitude (Fig. \ref{fig:lat}), and air humidity \citep{Laube:2014b}".  We have also tried to present  some of the complexities in our new Figure 1. We note, however, that our results are strong (e.g., 95\% uncertainty intervals for estimates of chilling, photoperiod, and forcing do not overlap zero) and suggest that, despite these complexities, consistent effects emerge. \\
\par (No third point listed by the reviewer)
\emph{4) species-specific response to chilling, photoperiod and forcing. This has been well reported, for example the pioneer species are opportunistic and photoperiod-insensitive, in contrast the late successional species are sensitive to photoperiod and higher forcing requirements, see the papers, as the authors cited, Körner \& Basler 2010;2014; Laube et al, 2014; Zohner et al, 2016 and other studies. Across so large dataset/many species, the mean values, for example chilling effect is 2 times larger than forcing and photoperiod as well as its sensitivity, hold large uncertainty and are no sense.}\\

\par We completely agree with the reviewer that species-specific differences are important. Indeed our modeling approach is designed to help examine across-species effects but also species-level differences. We now clarify this on Lines 62-65, where we write ``... \\

\emph{One of the main conclusions is that chilling is over-weight forcing and recent advanced leaf-out is mainly associated with spring warming. However, this is inconsistent with recent study that found the spring phenology did not significantly change during the global warming hiatus, see its figure 1 in Wang et al, 2019 Nature comm, but the spring T is still significantly increase and winter getting colder over the Eurasian (Li, Stevens and Marotzke 2015 GRL)). It seems that increasing chilling and forcing could not explain the dynamics in spring phenology? How to explain this inconsistency?}\\

% EMW: So far, all replies seem good, below is good, we could maybe add a little language to main text about complexities to also address this. 
\par We thank the reviewer for thinking critically about how our paper may contrast with previous work, and seeking to understanding the implications of our work. We now cite these two papers in Lines XX and  We believe the reviewer is suggesting that, if winters got cooler (and chilling therefore increases), as the Reviewer suggests  Li, Stevens and Marotzke (2015) find, one might expect spring phenology to  advance, given our hypothesis and findings. However, this is not what was found according to Wang et al (2019). We wish to point out to the reviewer that a decrease in temperature is not necessarily equivalent to an increase in temperature, given the thresholds involved in estimating chilling. For example, in this paper we find that warming actually increases chilling in many locations (Supp Fig). Thus, cooling might be decreasing chilling estimates and could sto; be consistent with the Wang et al paper, as well as our findings. \\

\emph{Minor commons}\\
\emph{Line numbers are needed;}\\
\par We thank the reviewer for this suggestion and have added line numbers. \\

\emph{In methods, the study yielded data from 72 studies across 39 yrs... this is misleading, because for many experimental studies, table S1, the data only for one year, and most less than 3 yrs.} \\

We thank the reviewer for pointing this out and can see how it was unclear. To address this concern, we have removed reference to the 39 years and added that `each experiment generally lasts under a year' to line XX. \\

\emph{More description is needed of Bayesian hierarchical model in the main text;} \\ 

\par We thank the reviewer for pointing out the need for more detail on our model. We now describe the model in more detail and include the equation for our main budburst model (Lines 61-77). \\

\emph{In the results sections, chilling has greater effect on budburst than forcing?. I would suggest providing the conditions, i.e. under future climate warming, due to the fact that these results come from experimental studies that simulated future warming, }\\

\par We have added a phrase to clarify that the scope of our results applies to controlled environment studies by saying (Lines 95-97): ``Our results, however, suggest that, across 203 species and 72 controlled environment studies around the world, chilling has a greater effect on budburst than forcing" \\ %EMW: We've hcanged a bit regarding this now.

\emph{In the results sections as well, the chilling only occur at warming above 4°C? interesting, but does it occur across species? and locations?} \\

\par We thank the reviewer calling our attention to the near for more detail and nuance here. We have modified the sentence so that it now says (Lines XX-XX): ``..our results suggest that delays due to decreased chilling only occur at warming above at least 4\degree C for most sites, though responses vary by species (Fig. 4, S6, S9).'' \\

\emph{Zohner, Constantin M., et al. ``Rising air humidity during spring does not trigger leaf-out timing in temperate woody plants." New Phytologist (2019).
Wang, Xufeng, et al. ``No trends in spring and autumn phenology during the global warming hiatus." Nature communications 10.1 (2019): 2389.
Li, Chao, Bjorn Stevens, and Jochem Marotzke. "Eurasian winter cooling in the warming hiatus of 1998-2012." Geophysical Research Letters 42.19 (2015): 8131-8139.}
-cite these in main ms
-could add humidity when to main text- \\


 \emph{Reviewer \#3 (Remarks to the Author):}\\

\emph{This manuscript addresses the relative importance of the environmental determinants of plant phenology using a meta-analytical approach. Specifically, the authors combine the experimental results of 72 studies and 203 species to estimate the effects of day length, winter chilling, and forcing on spring phenology, using hierarchical Bayesian models. The main finding is that almost all species respond to all three cues, with chilling having the largest, day length the smallest effect. Furthermore, the results suggest that, while all cues are important under experimental conditions, spring forcing will remain the dominant driver of spring phenology over the coming decades. The manuscript is well written and addresses a clear question. However, I have reservations as to the overall importance and validity of these results. That chilling is more important than day length has been shown by previous multi-species studies addressing this (e.g., Laube et al. 2014, Zohner et al. 2016).}\\

We thank the reviewer for his/her review and appreciate s/he found the manuscript well-written and addressing a clear question. We fully agree with the reviewer that other studies have found chilling is more important than daylength, but other studies have stressed the importance of photoperiod effects \citep[e.g.,]{vitasse2013,Basler:2014aa} and our study is one of the very few to provide a comparison of all three cues. Given the current debates in the literature over the relative importance of these cues we feel a meta-analytic approach---as we present here---is critical to advancing research. Further, our manuscript and related database provide a way to begin to understand why studies may differ: for example, we find that studies using sequential removal of tissue from the field yield different estimates than studies that manipulate chilling in controlled environments (lines XX). We thus believe our manuscript provides firmer evidence for some trends, provides a unique cross-study comparison of all cues and advances research through new efforts to understand why studies may differ in their findings. \\

\emph{Furthermore, the model output seems to suggest that all three cues (day length, chilling, and forcing) affect phenology in almost all species, leading the authors to conclude that their results contrast with the extensive literature [Zohner et al. 2016, K\"{o}rner \& Basler 2010*] suggesting photoperiod is an unimportant cue for many species. [page 4]? Yet, when looking at Table S2, most of the species-level data they use are taken from Zohner et al. (2016) [Zohner16 database]. In fact, 173 (85\%) of the 203 species included in this study were already investigated in Zohner et al. (2016). Given that in Zohner et al. (2016), 112 (65\%) out of 173 studied species did not react to daylength at all, it is surprising that day length is reported as a relevant, consistent cue across species. This makes me wonder whether their hierarchical Bayesian model is confounded (e.g., giving to much weight to certain species ?complexes?) and thus not suitable for exploring the relative importance of the different environmental drivers of spring phenology.}\\

The reviewer raises a good point, which we believe overlaps in part with Reviewer 1's concerns over clarifying how we handled species and studies in our models. We have now completely overhauled our explanation of this in the main text to clarify that we focus on results from a model of only 37 taxa that are well-represented across studies and designs. We did this to deal with the issue the reviewer is raising---in a meta-analytic approach such as ours we do not want our results to be dominated by any one study. Instead the aim is to present a quantitative synthesis of the literature. In this model, data from \citet{zohner16} make up only 8\% of the total data, which may explain in part why we found different results from \citet{zohner16}. To rigorously examine whether \citet{zohner16} affected the estimates in our model, we refit the model, excluding all daa from \citet{zohner16}. Estimates of chilling, forcing, and photoperiod were qualitatively consistent with model estimates from the full dataset. We summarize this model in Table SX in the Supplemental Materials.

\emph{K\"{o}rner \& Basler 2010 clearly is an inadequate reference here, please delete}\\

\par We have removed this reference from the manuscript in all locations used and replaced it with other references that include more primary research data. Additionally, in our original submission, we cited two papers by K\"{o}rner \& Basler 2010: a `Perspective' in \emph{Science} (K\"{o}rner \& Basler 2010,  \emph{Science} 327, 1461) and a response to a critique of this perspective (K\"{o}rner \& Basler 2010, \emph{Science} 329, 278)---we now have removed and replaced both references from our manuscript. \\

\emph{Apart from that, I take issue with the estimation of the importance of forcing and the attempt to estimate the relative importance of day length, chilling, and forcing. First, I don't see how the effect of forcing can be disentangled from the effects of chilling. This would require knowledge on which temperature ranges are adequate to satisfy chilling and forcing requirements. Yet, as correctly stated in the Supplementary information (page 2), current models of chilling are hypotheses and likely to be inaccurate for many species. Similarly, the effective temperature ranges to fulfill forcing requirements are not known. As such, when comparing the relative importance of winter chilling versus spring warming both factors are likely to be confounded. Also, if a study uses two different forcing temperatures that both lie within the range of optimal forcing conditions, one would see no effect between the treatments and the authors would thus infer that forcing didn't affect phenology, when in fact, forcing has a huge effect, not detected by the study design. Given these considerations, I don't think that a multivariate model, such as the one presented in this study, can adequately disentangle the relative importance of the three main phenological cues. }

\par The reviewer makes an astute and important point that it is impossible to fully disentangle effects of all three cues in many experiments, given the way that these treatments are applied. We completely agree with the reviewer's point and have added a new figure and adjusted our text throughout the manuscript (abstract, main text, conclusions) to address this. We added a new Figure 1 in this version which lays out some of these challenges explicitly for readers. As the reviewer points out, there is a great need for specifics-specific information about the effective temperature ranges for fulfilling chilling requirements. One motivation for this paper is to highlight the need for additional work on this and other aspects of spring phenology. We have also added a new sentence to the abstract, where we write, ``Major progress beyond this will require greater efforts to fully separate chilling and forcing effects at the physiological-level.''\\

\par In absence of these species-specific data and given our meta-analytic approach, we rely on the original researchers to separate forcing from chilling conditions. Our database  reports the treatments imposed by these researchers. Therefore, an assumption underlying out work is that they used a range of treatments that are relevant for their focal species. We now state this explicitly in Lines XX, where we write..."" \\
% Changes made:
% New figure
% New sentence in abstract
% Supplementary material
% EMW: R3 is also complaining about threshold effects of forcing, we could sneak in a clause about this in our chilling models paragraph ... though I personally think forcing is more linear. We could at least add something about this to the supp? We just need to make sure we don't skip the point ... (see below comment)
 
% EMW: Seems good -- main reply should be: above, plus maybe add some text about threshold/optimum effects [reviewer's point that 'if a study uses two different forcing temperatures that both lie within the range of optimal forcing conditions, one would see no effect between the treatments and the authors would thus infer that forcing didn?t affect phenology, when in fact, forcing has a huge effect, not detected by the study design'] and our how our models try to handle this. [Can we also cite some figures in the supp for this? I thought this was an issue but I think we actually did see a effect most often.]

-cite new schematic figure
- one of the findings of our paper- is that we need more work to accurately separate chilling and forcing
- the perceptive reviewer has highlihgted a general problem of the field...and yet...our models has been exrtemely predictive. 
\\

\emph{p.2: What do you mean by `we included only studies with at least 49.5\% budburst?' This is not correct for most of the studies included in your OSPREE dataset. E.g., Heide (1993) and Zohner et al. (2016) defined budburst as the date when 1-3 buds on a twig had opened. Please clarify.}\\

\par This is a good point and differences in the definition of budburst are exactly why we have this \emph{Defining budburst} section. We have now clarified our meaning. What now have adjusted text on lines xx-xx to read as follows, "." This issue was only for studies that reported percentage of budburst as a function of time; to make these studies comparable with most other studies, we needed to estimate one day of budburst for each treatment from the \% budburst by time figures (this was a limited number of studies). In general, we extracted a date close to 90\% for most studies of this type,  but our code allowed studies to be included as long as the highest \% budburst was at least 49.5\%. We realize this is confusing and hope we have clarified this point now. \\

\emph {p. 3: Total chilling ranged from -1304 to 4724 Utah units? The Utah model allows for negative chilling units? What's the biological justification for that?}\\

\par We agree with the reviewer that the Utah model, as well as other chilling models, are often non-intuitive and biological justifications may not be immediately apparent. We now describe this model, as well as the other chilling model we used, Chill portions, on Lines XX  in the main manuscript. We used the Utah model to report chilling because that allowed us to include the greatest amount of studies from the OSPREE database (i.e., because many studies used this model to estimate chilling).  

\par The biological justification, as described by the developer of the original model (Richardson) is ... \\%AKE: waiting for this article still- can't find it in my files!

\par We have also tried to make clear that in the main text that all chilling models are hypotheses on Lines 112-116, where we state:  ``Models of how species accumulate chilling are poorly developed for forest trees, with few relevant tests evaluating the particular temperatures at which species do or do not accumulate chilling. Instead, researchers generally rely on models developed for perennial fruit trees (i.e., Utah (32) and chill portions (37), both of which were developed for peach species). These models are themselves hypotheses for how chilling may accumulate and produce dormancy release, but are likely to be inaccurate for many species (33).These models are themselves hypotheses for how chilling may accumulate and produce dormancy release, but are likely to be inaccurate for many species (33)."

\emph{p. 4: Latitude model: This model doesn't make sense to me. What is the latitude you refer to here? The location where the experiment took place? You refer to provenance locations, I doubt these are available for most of the studies, especially the ones conducted in botanical gardens or other collections.}\\

\par We thank the reviewer for highlighting the need for greater clarity about our terminology and the model. We now clarify our meaning and have adjusted the model to include only studies that give locations of the populations they sampled from. In the initial submission, we did not adequately clarify what was meant by latitude: we now refer to this as `provenance latitude,' throughout out the manuscript (see below for specific changes by line number). We also revisited all papers used in this model to check that they refer to the latitude of origin (geogragraphical source of the population) of plant material used in the experiment, which the reviewer is correct in noting is the traditional definition of provenance. Most studies did give clear provenance, and many were actually studies of multiple provenances \citep[e.g.,][]{Heide:1993a,Falusi:1996aa,Partanen:1998aa,Li:2005aa,Basler:2012}. We found, however, that several studies did not include provenance location, were crop species (e.g., commercial blueberries, rather than wild blueberries), or were of unclear provenance \citep[e.g.,][]{zohner16}, which we had intended to remove from this analysis. We refit the latitude model to this slightly smaller group of species (a less than 15\% reduction in data) and found that estimates of chilling, forcing, photoperiod, and latitude remain qualitatively consistent with our previous model. We now present this more restrictive model in our manuscript and thank the reviewer for catching this discrepancy. 
\begin{enumerate}
\item Lines 93-95 in the main text: we now say `While photoperiod had the smallest effect among the three cues, our results contrast with the extensive literature suggesting photoperiod is an unimportant cue for many species (7; 5). |instead we found it was surprisingly large, even when accounting for its interaction with provenance latitude (i.e., the latitude of origin of plant material; see Supplemental Materials for details, especially Figs. S3, S11, Table S5)."
\item Supplemental Materials, Page 4: in the description of the Latitude Model, we now say: 'we examined the effect of including provenance latitude in a model similar to our main one, but designed to estimate effects of provenance latitude. This model estimated the effects of each phenological cue (chilling, forcing, photoperiod) on days to budburst (as in the main model), in addition to the effect of provenance latitude (i..e., the latitude of origin of plant material used in the experiment) and the interaction of photoperiod and provenance latitude. We include this interaction because photoperiod effects are expected to vary by latitude...' and in the next paragraph we now say `..then subsetted the species and species complexes to include only those that had multiple provenance locations across different latitudes.'
\item Supplemental Materials, Caption for Table S5: We have replaced 'latitude' with 'provenance latitude' so that the caption now reads: `Using a model with Utah chilling units and testing the effects of provenance latitude plus the interaction between provenance latitude and photoperiod results in slightly muted effects...'
\item  Supplemental Materials, Caption for Figure S3: We have replaced 'latitude' with 'provenance latitude' so that the title of the caption now reads: `Estimates for effects of chilling exceeded estimates for forcing, photoperiod, provenance latitude, and the interaction between latitude and photoperiod, for most species...'
\end{enumerate}

\emph{Figures: Figs. 2 and 3, showing a 3-dimensional illustration of the interplay between winter chilling and spring warming, are very hard to read. I would prefer a simpler illustration.}\\
\par We appreciate the Reviewer's perspective, and struggled ourselves with differing preferences among co-authors and reviewers. We have modified Figure 2 so that it includes both a simpler 2-dimensional version and  the previous 3-dimensional version. We have shown both simpler 2-dimensional versions of these figures and the more complex 3-dimensional versions to a number of scientists and found opinions to be split on preferences for 2d versus 3d versions. We hope that showing both versions in Figure 2 facilitates the reader's understanding and interpretation of the 3-dimensional illustrations. 



%Thoughts from Inaki:
%- Maybe we should think less of chill being 3X more important than forcing but that this shows the hierarchy -- C >> F >> P.
%- Sequential model means the ONLY dependency across parameters across models is that one phase must finish before the next.
%- Overlapping or alternating or all parallel: are common other models -- Described in Chuine 2000 (Phenofit) paper -- they all say that amount of forcing depends on amount of chilling (alternating also covered in Pope et al. 2014). 
%- Alternating shows up clearly in biological settings but is hard to explain at a physiological level. He thinks the 'growth competence' (aka parallel) model produces the same effect but is biological more realistic (it's in the Caffarra Betula papers). It suggests that the time to truly transition out of and fully release endodormancy can be fast or slow and depends on these factors. 

\bibliography{..//..//..//..//refs/ospreebibplus.bib}

\end{document}
